---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r, message=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(car)
bird_data <- read_table("data/britain_species.txt", col_names = FALSE)
```


```{r}
bird <- bird_data %>% 
  rename(Island = X1,
         Area = X2,
         Elevation = X3,
         Soil = X4,
         Latitude = X5,
         Distance = X6,
         Species = X7) %>%
  mutate(Area = as.double(Area),
         Elevation = as.integer(Elevation),
         Soil = as.integer(Soil),
         Latitude = as.double(Latitude),
         Distance = as.double(Distance),
         Species = as.integer(Species))
head(bird)
```


```{r}
bird.reg <- lm(Species ~ Area + Elevation + Soil + Latitude + Distance, data = bird)
```

## Collinearity
```{r}
temp_bird <- bird %>% 
  select(-Island)

cor(temp_bird)
```


## Multicollinearity
```{r}
vif(bird.reg)
```

# Model Selection by Stepwise Regression

```{r}
step(bird.reg, direction = "both")
```


# Partial F test Full/Reduced Model

H0: Beta1 (Area) = Beta2 (Elevation) = 0 | Ha: ALOI | Alpha = 0.05


```{r}
bird.reg1 <- lm(Species ~ Soil + Latitude + Distance, data = bird)

selcri <- function(lmout){
  n <- length(lmout$fit)
  rsq <- summary(lmout)$r.sq
  adj.rsq <- summary(lmout)$adj.r.sq
  aic <- extractAIC(lmout)[2]
  bic <- extractAIC(lmout, k = log(n))[2]
  press <- sum((lmout$residuals/(1 - hatvalues(lmout)))^2)
  cbind(rsq, adj.rsq, aic, bic, press)
}
anova(bird.reg1, bird.reg)
```

Because the p value = 0.4896 > alpha level = 0.05, we fail to reject the null hypothesis and chose the reduced model by dropping the Area and Elevation variables from the model.

```{r}
selcri(bird.reg)
selcri(bird.reg1)
```

The Reduced Model statistically fits the data better because the adjusted R^2 is higher indicating higher correlation for the reduced model, and the AIC, BIC, and PRESS values are lower indicating there is less error

## Residuals
```{r}
plot(bird.reg1$residuals ~ bird$Species)
abline(0,0)
```


## QQPlot
```{r}
qqPlot(bird.reg1)
```

# Full/Reduced Model without Britain

Hypothesis: Because Britain is such a large outlier in terms of its Area and Elevation compared to the other British isles, then if it were removed the new models would fit the data better

```{r}
NOBritainBird <- bird %>%
  filter(Island != "Britain")
NOBritainBird.reg <- lm(Species ~ Area + Elevation + Soil + Latitude + Distance, data = NOBritainBird)
step(NOBritainBird.reg, direction = "both")
```


```{r}
NOBritainBird.reg1 <- lm(Species ~ Soil + Latitude + Distance, data = NOBritainBird)
anova(NOBritainBird.reg1, NOBritainBird.reg)
```

H0: Beta1 (Area) = Beta2 (Elevation) = 0 | Ha: ALOI | Alpha = 0.05

Because the p value = 0.3466 > alpha level = 0.05, we fail to reject the null hypothesis and chose the reduced model by dropping the Area and Elevation variables from the model.

```{r}
selcri(NOBritainBird.reg)
selcri(NOBritainBird.reg1)
selcri(bird.reg1)
```

```{r}
influenceIndexPlot(NOBritainBird.reg1)
```


What can we conclude from this?






