---
title: "Bird_Analysis"
author: "Reagan Edgren"
date: "12/6/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(GGally)
library(ggpubr)
```



```{r}
bird_data <- read_table("~/GitHub/stat-415-final-project/data/britain_species.txt", col_names = FALSE)

#Modified path
#bird_data <- read_table("data/britain_species.txt", col_names = FALSE)

bird_data %>% 
  rename(island = X1,
         area = X2,
         elevation = X3,
         soil = X4,
         latitude = X5,
         distance = X6,
         species = X7) %>%
  mutate(area = as.double(area),
         elevation = as.integer(elevation),
         soil = as.integer(soil),
         latitude = as.double(latitude),
         distance = as.double(distance),
         species = as.integer(species)) -> bird_data
bird_data
```


Estimated regression function
```{r}
bird.reg <- lm(species ~ area + elevation + soil + latitude + distance, 
               data = bird_data)

summary(bird.reg)
anova(bird.reg)
```

```{r}
plot(bird.reg$residuals ~ bird_data$species)
abline(0,0)
```

correlation regression

```{r}
bird_data %>%
  select(area,elevation,soil,distance,species) -> bird_cor
cor(bird_cor, use = "complete.obs")
```

```{r}



# ggarrange(area1, area2, elev, soil, dis, spe, lat,
#           ncol = 2, nrow = 4)
bird_data %>%
  select(species, area, elevation, distance, soil, latitude) %>%
  ggpairs()

```




can we get rid of latitude (partial F)

```{r}
bird_nolat <- lm(species ~ area + elevation + soil + distance, 
               data = bird_data)

anova(bird_nolat, bird.reg)
```

```{r}
bird_noarea <- lm(species ~ latitude + elevation + soil + distance, 
               data = bird_data)

anova(bird_noarea, bird.reg)

bird_noelevation <- lm(species ~ latitude + area + soil + distance, 
               data = bird_data)

anova(bird_noelevation, bird.reg)

bird_nodistance <- lm(species ~ latitude + elevation + soil + area, 
               data = bird_data)

anova(bird_nodistance, bird.reg)
```

```{r}
bird_nolat <- lm(species ~ area + elevation + soil + distance, 
               data = bird_data)

anova(bird_nolat, bird.reg)

bird_nosoil <- lm(species ~ latitude + elevation + area + distance, 
               data = bird_data)

anova(bird_nosoil, bird.reg)
```




```{r}
bird_no <- lm(species ~ latitude + soil, 
               data = bird_data)

anova(bird_no, bird.reg)
```

# Step-wise regression
Slight modifications to formatting
```{r}
bird <- bird_data %>% 
  rename(Island = X1,
         Area = X2,
         Elevation = X3,
         Soil = X4,
         Latitude = X5,
         Distance = X6,
         Species = X7) %>%
  mutate(Area = as.double(Area),
         Elevation = as.integer(Elevation),
         Soil = as.integer(Soil),
         Latitude = as.double(Latitude),
         Distance = as.double(Distance),
         Species = as.integer(Species))
head(bird)
```


## Collinearity
Before we delve into the role 
```{r}
temp_bird <- bird %>% 
  select(-Island)

cor(temp_bird)
```


## Multicollinearity
```{r}
vif(bird.reg)
```

# Model Selection by Stepwise Regression

```{r}
step(bird.reg, direction = "both")
```


# Partial F test Full/Reduced Model

H0: Beta1 (Area) = Beta2 (Elevation) = 0 | Ha: ALOI | Alpha = 0.05


```{r}
bird.reg1 <- lm(Species ~ Soil + Latitude + Distance, data = bird)

selcri <- function(lmout){
  n <- length(lmout$fit)
  rsq <- summary(lmout)$r.sq
  adj.rsq <- summary(lmout)$adj.r.sq
  aic <- extractAIC(lmout)[2]
  bic <- extractAIC(lmout, k = log(n))[2]
  press <- sum((lmout$residuals/(1 - hatvalues(lmout)))^2)
  cbind(rsq, adj.rsq, aic, bic, press)
}
anova(bird.reg1, bird.reg)
```

Because the p value = 0.4896 > alpha level = 0.05, we fail to reject the null hypothesis and chose the reduced model by dropping the Area and Elevation variables from the model.

```{r}
selcri(bird.reg)
selcri(bird.reg1)
```

The Reduced Model statistically fits the data better because the adjusted R^2 is higher indicating higher correlation for the reduced model, and the AIC, BIC, and PRESS values are lower indicating there is less error

## Residuals
```{r}
plot(bird.reg1$residuals ~ bird$Species)
abline(0,0)
```


## QQPlot
```{r}
qqPlot(bird.reg1)
```
